---
title: "Homework 1"
subtitle: "Soc 225: Data & Society"
author: "PUT YOUR NAME HERE"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

Write all code in the chunks provided. Complete this `.Rmd` file and knit it into an `.html`. You must upload both files for credit. 

Remember to unzip to a real directory before running everything!

# 1. Basics of R and R markdown

## 1.1. Create a vector containing elements 10, 22, 27, 19, 20 and assign it with a name. 

```{r}

```

## 1.2. Use R as a calculator to compute the following values. 

a) 27(38-17)
b) ln(14^7)
c) sqrt(436/12)

```{r}

```


## 1.3. Run the below code to create a vector. Observe what e contains and use `?seq` to see help of function `seq()`.
```{r}
e <- seq(0, 10, length=5)
e
```

Create the following vectors:
  b = (87, 86, 85, ..., 56)
  
```{r}

```

What is the 19th, 20th, and 21st elements of b?

```{r}

```


## 1.4. Compute the following statistics of b:
  a) sum
  b) median
  c) standard deviation

## 1.5. Following the example given in lab1, mix in-line R calculations with text and make reference to vector b. You must use in-line R calculations at least once (e.g. functions like mean(), sd(), max()) and *may not hard-code any numbers referenced in your text*. An example is given below:

The average of `b` is `r mean(b)`.

# 2. Research Question (You don't need code for this question)

*For this problem you'll answer some questions to help explore your interests in data science. These are questions that you're interested in. They don't have to be things that you know the answer to and still less new areas of study.*  
  
*However, problem 3 asks you to come up with a 'big data' dataset that you think you might use to answer your question. If you're new to R or not sure about what to do, I encourage you to use the Airbnb data that we'll be using in class. In that case, make sure that your answers to problem 2 relate to the airbnb data.*

## 2.1: What are some areas of interest for you within sociology, big data, and computational social science?

## 2.2: Provide a link to a dataset which you think intersects with one of your interests. Explain the connection. You can find datasets by doing a google search or by looking here:http://hadoopilluminated.com/hadoop_illuminated/Public_Bigdata_Sets.html or here: https://www.kaggle.com/

If you're using the airbnb set, explain how it connects to your interests.

# 3. Import data and identify variables

## 3.1. Import your data into R and output the column names.

```{r}

```

## 3.2. Use View(), head() or tail() to check your data. What variables does it contain? How many rows are in your data? What is the unit of analysis in your data?

```{r}

```

## 3.3. Discuss how might some variables serve your research interest as discussed in problem 2 above. 


