---
title: "Homework 2"
subtitle: "Soc 225: Data & Society"
author: "PUT YOUR NAME HERE"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

Write all code in the chunks provided. Complete this `.Rmd` file and knit it into an `.html`. You must upload the `.html` files for credit. 

Remember to unzip to a real directory before running everything!

Problem 1 should be roughly analogous to what we've done in class, with a few extensions. There are hints at the bottom of this document if you get stuck. If you still can't figure it out, go to google/stack exchange/ask a friend. Finally, email your TA or come to office hours :).

```{r,message=FALSE,warning=FALSE}
## Run this code, to manage packages and install as needed.
# Install pacman
if (!require("pacman")) install.packages("pacman", repos = "http://cran.us.r-project.org")
# p_load function loads packages if installed, or install then loads otherwise
pacman::p_load(tidyverse)
```

# Problem 1: Piping Hot Variables

*This problem uses `dplyr` verbs to answer questions about an Airbnb data set.* 

## 1.1: Get the data

Go to [Inside Airbnb](http://insideairbnb.com/get-the-data.html) and download the "Detailed Listings" data for Seattle, `listings.csv.gz`. This file has many more variables than the "Summary" file we've been using in class. Put it in a `data/` subfolder in your `hw-02` project folder. 

[This is a compressed (gzipped) file, but R should be able to handle it as-is. If you run into trouble, try unzipping the file before reading it into R.]

## 1.2: Set up your R environment 

a. Load the tidyverse
b. Read the detailed Airbnb data into R
  
```{r}

```
  
## 1.3: Use the data to answer a question

*For how many units does the host live in a different neighborhood from the listing? For how many units does the host live in the same neighborhood as the listing?*

Try to figure out which variables to use from their names, and think about which verbs you've learned about might work to answer this question. See the hints at the end if you need help. 

```{r}

```

## 1.4: Build on your answer
  
*Building on that work, what is the average number of listings for hosts that live in the same neighborhood as their listing? What's the average for hosts who live in different neighborhoods from their listing?* 

The `mean` function will take the average of a variable, but you might need to look up how to use it. See the hints for more suggestions if you get stuck.

```{r}

```
  
## 1.5: Reflect and interpret

Reflect on your answer to 1.4. What might cause the results you got? How does that connect to the idea that Airbnb might be changing neighborhoods?  

*Your answer should be at least a few sentences here*

# 2. Prepare and Visualize data

## 2.1. Set up your environment

Set up your environment by: 

Reading the Airbnb data: There's another new data set in the `data/` folder. This one has almost 10,000 cases and the census data by zipcode. These data are from New York City, not Seattle! 

```{r}

```
  
We've given you absolute populations and proportions for the racial composition of the zipcode for each listing. We've also made a variable called 'modal_race' which is the race with the largest proportion in that neighborhood. 

These variables are all in the last columns of the data set---you can try selecting them and using `summary()` to get a sense for what they contain.

## 2.2: Turn `price` into a number

`price` includes dollar signs, which means that R interprets it as a character. We want it to be a numeric variable instead. Turn `price` into a numeric variable in the chunk below.

There are a few ways to do this using `tidyverse` functions. See the hints below for some suggestions.

```{r}

```

## 2.3: Make a scatterplot

Use a scatter plot to compare how unit prices change with the proportion of a particular race. 

Bonus: try grouping by zipcode (in any fashion) for this plot

```{r}

```
  

## 2.4: Make a boxplot
  
Use the `modal_race` variable to plot a boxplot comparing race and price. You may have to look up how to make a boxplot in `ggplot2`---what geom do you need?

Bonus: try showing how this comparison differs by neighborhood group.

```{r}

```

## 2.5: Interpret your answer

Interpret your answer to 2.4. Check the hints if you need help.

*Your answer should be at least a few sentences here*

## Bonus: how did we make the data?

There's another file in the data folder, census.csv. Read it into R and have a look at it.

Download the full listings for New York City from Inside Airbnb, and see if you can join the Census data to it by zipcode using `left_join`. You'll have to filter out some weird values for zipcode before you can merge. 

```{r}

```


# 3. Your own data

## 3.1. Looking at the dataset you chose for homework 1, think about a research question you'd like to investigate (try search about existing studies around your question). What variables do you plan to use to answer your question?

## 3.2. What is one way that you have to modify or examine your data to begin to answer your question?

## 3.3. Using the functions we've worked with in class (select, filter, arrange, mutate), plus any others you'd like to use, clean and transform your data set to make it ready for further exploration.

You must:

a. Create a new dataset that only includes the variables you're interested in
b. Output a version of that dataset that only includes certain values of observations, hopefully ones you're interested in.
c. Order your data by the values of one variable you're interested in. 
d. Create a modified version of one of your variables (many of you will *need* to do this, but even if you don't, I want to see that you can)
e. Look up and try out one new verb for data transformation. The RStudio data transformation cheat sheet is a fantastic place to start: https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf

For e., we'd recommend using `group_by` + `summarize`. You can group your data by one variable, and then see the mean (or similar) of another variable within each of those groups.

*Use as many code blocks as you need for a-e*
  
```{r}

```

# Hints
1.3 Try using these steps:

- Step 1: identify the variables you need
  - Listing neighborhood: `neighbourhood`
  - Host's neighborhood: `host_neighbourhood`

- Step 2: Filter the data to only include the rows where those variables are not equal. Look back to Module 2 (or look online) if you need a reminder about how to write "equal", "not equal", and so on in R.

- Step 3: How many rows are left in the filtered data?

Extra food for thought: how do "NA" (missing) values get handled here? Do you think that makes sense? Should you do something else with them, maybe using `is.na`?

1.4 The variable for number of listings is `host_listings_count`. You might want to make a new variable indicating if a host is a local host (your answer to 1.3 will help here!). There are many ways to use `mean` on a subset of data, but the best approach is one we introduce in Module 5: `group_by` + `summarize`. Try it out now if you can! For this problem, don't worry about NAs. 

2.2

Use `mutate` for this. You can replace the original `price` variable, or name it something else. There are a couple things you can use on `price` inside the mutate: 

- `parse_number`, a function in the `readr` package, does a good job of converting currency to numbers on its own.

- `str_extract` with `pattern = "\\d+"`, then `as.numeric`, will extract numbers from a string, then convert the new (sub)string to a number.

- `str_remove_all`, with `pattern = "[\\$|,]"`, then `as.numeric`, will remove all dollar signs and commas.


2.5

Check out these resources if you're not sure about interpreting box plots:

https://magoosh.com/statistics/reading-interpreting-box-plots/

https://www.youtube.com/watch?v=oBREri10ZHk 

3.3

a. use select()
b. use filter()
c. use arrange()
d. use mutate()
e. use group_by(var1) %>% summarise(mean = mean(var2))
