---
title: "Homework 3"
subtitle: "Soc 225: Data & Society"
author: "PUT YOUR NAME HERE"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

Write all code in the chunks provided!

Remember to unzip to a real directory before running everything!

Problems should be roughly analogous to what we've done in class, with a few extensions. There are hints at the bottom of this document if you get stuck. If you still can't figure it out, go to google/stack exchange/ask a friend. Finally, email your TA or come to office hours :).

# Problem 1: Google Trends

## 1.1 
*Go to Google Trends and search for "covid-19 vaccine". Look at variations by time and by region in US. What do you observe?*

# Problem 2: Join data frames

In this problem we will use data in the `nycflightdata13` package to perform joining of data frames.

It includes five dataframes, some of which contain missing data (`NA`).

- `flights`: flights leaving JFK, LGA or EWR in 2013
- `airlines`: airline abbreviations
- `airports`: airport metadata
- `planes`: airplane metadata
- `weather`: hourly weather data from JFK, LGA and EWR

Note these are **separate data frames**, each needing to be loaded separately using `data()`.

## 2.1. Set up your environment:

a. Install and load the `nycflights13` package. Load the `tidyverse` package.
b. Load data sets `flights`, `planes`, `airlines`

```{r}

```

```{r}

```

## 2.2 Find data frames
*We'll be looking at who manufactures the planes that flew to Seattle. Which are the two data frames we need to join?*

## 2.3. Find common keys
*Take a look at variables contained the two data frames. Which variable(s) should be used as the key to join?*

## 2.4. Join the two data frames

```{r}

```

## 2.5. Build on your answer
*For flights with a destination of Seattle, who are the largest manufacturers? Give top five of the manufacturers.*
(Check hints if you have troubles)
```{r}

```

## 2.6. Use the data to anwer the below questions
*We'd like to know which airlines had the most flights to Seattle from NYC. Which are the two data frames we need to join, and on which key variable(s)?*

```{r}

```

##2.7. 

*Join the two data frames in 2.6 and list the top five airlines.*

```{r}

```


# Problem 3: Your research question

Think about the research question you have in mind. Plot is a great way to understand patterns, key relationships and uncertainties in a data set. Here we'll ask you to plan about plotting your variables of interest for your research question. Try to think about **3 plots** below:

*For each of the 3 plots, provide:*
  
A. The purpose of the plot: what do you want people to understand when they see this?  

B. The type of plot: what geom functions will you use to present the plot? Why are those the best choices?  

C. Limitations/biases: What is missing from this presentation? Could someone get the wrong idea? What can you do to help limit the negative possibilities here?  

### Plot idea 1 

A.  
B.  
C.  

### Plot idea 2 

A.  
B.  
C.  

### Plot idea 3 

A.  
B.  
C.  


Hint

2.5

use `left_join()` by "tailnum" to join the two data frames, then count() observations by manufacturer, and then use arrange() with descending order.

