---
title: "Lab6: Google Trend and Search"
subtitle: "Soc 225: Data & Society"
author: "[PUT YOUR NAME HERE]"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

# Agenda
- Google Trends
  - Single term
  - Comparing multiple terms
  - Mapping search data from Google Trends
  - Embedding a Google Trends widget
- Google Ngrams
- Challenge: Google Takeout

# Google Trends:

## 1. Single term

Go to Google Trends (https://trends.google.com/trends/?geo=US) and search for "crisis actors." Look at variation by time and by region. 

**Question 1.1: why is there a spike in mid-February?** Look at the related topics and queries for hints at the answer.


## 2. Comparison

Compare the relative popularity of the following social media platforms using Google Trends: Facebook, Twitter, Instagram, and YouTube. 

**Question 2.1. According to Google Trends, what is the order of popularity for these platforms?**

**Question 2.2. Do you think volume of Google searches is a good metric of social media use?**

**Question 2.3. Come up with a set of search terms that you think would make an interesting comparison, and describe your results here.**

## 3. Mapping search data from Google Trends

In this section we'll learn how to get and visualize Google search data with both time and geographical components using the packages `gtrendsR`. In this example, we'll look at search interest for named hurricanes that hit the U.S. mainland and then plot how often different states search for "guns".

**Question 3.1. Install and load package `gtrendsR`. Load package `tidyverse`.**

```{r}

```

Let's first look at how the impact of hurricanes Katrina (August 2005) and Harvey (August 2017) are reflected in how Americans have used these names as google search items over time. 

The `gprop` argument controls whether we want general web, news, image or Youtube searches. 

The `time` argument is set to "all" and will gather data between 2004 and the time the code is run. 

Focusing on searches made in the U.S., we'll set the geo argument to "US". 
```{r}
# hurricanes <- gtrends(c("katrina","harvey"), time = "all", gprop = "web", geo = c("US"))
```

**Question 3.2. Take a glimpse at `hurricanes`. What does `gtrends` function actually returns?**

We plot the output:
```{r}
# plot(hurricanes) + 
#   geom_line(size = 0.5) 
```
The above plot shows clear spikes around the time when Katrina and Harvey hit the US. To understand what is actually measured on the y-axis, have a look [here](https://support.google.com/trends/answer/4365533?hl=en). 

We could also plot more cyclical data using the same method. 

**Question 3.3. Make and plot ten years' worth of search data from 2008-01-01 to 2018-01-01 in US for "spring break" and "vacation".**

```{r}

```

**Question 3.4. What do you find about search data for "spring break" and "vacation"? Do you find concurrence between the searches of these two terms?**

Next, let's compare the amount of interest in guns for each US state. 

**Question 3.5. Create an object "guns" that fetches google web search of word "guns" in US for all time. Take a glimpse at `guns`.**
```{r}

```

We then create a data frame with class `tbl_df` out of `interest_by_region` under `guns`.

```{r}
# gunsInterestByRegion <- as_tibble(guns$interest_by_region)
```

Next, create an object `statesMap` that fetches an empty map for plotting our geographical data. 
```{r}
# statesMap = map_data("state")
```

We'll next join `gunsInterestByRegion` with `statesMap`. 

**Question 3.6. Which columns in these two data frames could be used as keys to join? Are the labels identical under these two columns?**


We change the capitalized state names to lowercase using `stringr`. 
```{r}
# gunsInterestByRegion <- gunsInterestByRegion %>% 
#  mutate(region = str_to_lower(location))
```

**Question 3.7. Join `gunsInterestByRegion` with `statesMap`. Name the new object `gunsMerged`. Take a glimpse.**


```{r}

```

**Question 3.8. Use `gunsMerged`, create a map that shows Google search interest for guns in each state using `ggplot` and `geom_polygon`. Fill the map with the log of `hits`. Properly name your map.**

```{r}
# I like to add comments to plan out my work

# create a ggplot object with x being longitude and y being latitude
# add a map layer, group and fill the map properly
# add a title to the map
# Make the visualization pretty and easy to read
```
**Question 3.9. What do you find about the search interests of `guns` in different states? How do you explain your findings?**

## 4. Embedding a Google Trends widget

You can embed a Google Trends widget as an interactive object, but the steps are a little tricky: 

1. In Google Trends, click the "Embed" button ("< >") and copy the code
2. Paste it into your Rmd file, and knit to html
3. Open the html in Chrome
4. Inspect the chart, and copy the "iframe" element
5. Paste *that* into your Rmd file, and delete the original script
6. Add a height to the iframe (e.g. height="450px")
7. Regenerate and reopen the html


The final result should look like this (open html in Chrome): 

<iframe id="trends-widget-1" src="https://trends.google.com:443/trends/embed/explore/TIMESERIES?req=%7B%22comparisonItem%22%3A%5B%7B%22keyword%22%3A%22crisis%20actors%22%2C%22geo%22%3A%22US%22%2C%22time%22%3A%22today%2012-m%22%7D%5D%2C%22category%22%3A0%2C%22property%22%3A%22%22%7D&amp;tz=420&amp;eq=q%3Dcrisis%2520actors%26geo%3DUS%26date%3Dtoday%2012-m" width="100%" frameborder="0" scrolling="0" height="450px"></iframe>

**Question 4.1. Below, embed one of the charts from your Google Trends search in the previous question.**

# 5. Google Ngrams
The Google Ngram Viewer draws on Google Books to show the use of terms over time. It has some interesting uses, especially in digital humanities, but with important caveats: https://www.wired.com/2015/10/pitfalls-of-studying-language-with-google-ngram/

(What's a n-gram? A sequence of one or more words.)

Try looking at references to different communications technologies, e.g., the Internet, TV, and radio. Do the results change if you make the searches case-insensitive?

<iframe name="ngram_chart" src="https://books.google.com/ngrams/interactive_chart?content=Internet%2CTV%2Cradio&case_insensitive=on&year_start=1800&year_end=2008&corpus=15&smoothing=3&share=&direct_url=t4%3B%2CInternet%3B%2Cc0%3B%2Cs0%3B%3BInternet%3B%2Cc0%3B%3Binternet%3B%2Cc0%3B%3BINTERNET%3B%2Cc0%3B.t4%3B%2CTV%3B%2Cc0%3B%2Cs0%3B%3BTV%3B%2Cc0%3B%3Btv%3B%2Cc0%3B%3BtV%3B%2Cc0%3B%3BTv%3B%2Cc0%3B.t4%3B%2Cradio%3B%2Cc0%3B%2Cs0%3B%3Bradio%3B%2Cc0%3B%3BRadio%3B%2Cc0%3B%3BRADIO%3B%2Cc0" width=900 height=500 marginwidth=0 marginheight=0 hspace=0 vspace=0 frameborder=0 scrolling=no></iframe>


**Question 5.1. Search for another set of terms of interest, and embed the results below.** (See this article for some examples: http://techland.time.com/2014/01/16/google-books-ngram-viewer/)

# 6. Challenge: Google Takeout

Download some of your own data from Google Takeout: https://takeout.google.com/settings/takeout

I recommend only picking one or two applications! Otherwise, your download file will be very large. Read this Twitter thread for some examples: https://twitter.com/iamdylancurran/status/977559925680467968

Look at your Google data, and describe the data structure. See if you can read some of the files into R.

# References
Peer Christensen, [Mapping search data from Google Trends in R](https://www.storybench.org/mapping-search-data-from-google-trends-in-r/)