---
title: "Homework 4"
subtitle: "Soc 225: Data & Society"
author: "PUT YOUR NAME HERE"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---
Write all code in the chunks provided!

Remember to unzip to a real directory before running everything!

In this homework we will conduct an exploratory data analysis using the `diamonds` data set included by the `tidyverse` package. 

Problems should be roughly analogous to what we've done in class, with a few extensions. There are hints at the bottom of this document if you get stuck. If you still can't figure it out, go to google/stack exchange/ask a friend. Finally, email your TA or come to office hours.

# Problem 1. Variation along a single variable

## 1.1. Set up your environment by:

a. Load the `tidyverse` package. 
b. Load the `diamonds` data set by using `data()`.

```{r}

```

## 1.2. Examine the `diamonds` data set
*Check how many rows and what variables the data set contains.*
```{r}

```

## 1.3. Select a variable and plot its variations
*Pay attention to the type of variable and choose an appropriate plot.*

```{r}

```

## 1.4. More plotting
*Create a new data frame that contains only diamonds smaller than 3 carat. Plot the distribution of `carat` in this new data frame using a histogram, setting a binwidth of 0.01.*

```{r}

```

## 1.5. Build on your plot
*Looking at the plot you created in problem 1.4, do you have any follow up questions about the data?*
(Check hints below if you have troubles)

# Problem 2. Covariation

## 2.1. A categorical and continuous variable
*Plot the distribution of price by cut. Which type of plot do you find most appropriate?*
(Check hints below if you have troubles)

```{r}

```

## 2.2. Two categorical variables
*Use `geom_count` to explore how diamonds color vary with cut. Are there other approaches you could think of to explore the question?*

```{r}

```

## 2.3. Two continous variables
*Visualize the covariation between the carat size and price of a diamond. What do you find?*
(Check hints below if you have troubles)
```{r}

```

## 2.4. More plotting
*Combine two of the techniques you've learned to visualize the combined distribution of cut, carat, and price.* 

```{r}

```

Problem 3. Your own data
Using your plans from Homework 3, produce the plots in R. Use as many code blocks as you need. Make sure to include proper titles and labels, and think about different ways to improve your plots. 

*What makes a chart good?* There are different takes on this: 

- Edward Tufte says you should think about reducing chart junk and improving your data/ink ratio.
- Zan Armstrong argues that what's "good" depends on your question, and you should make many different charts to see different things.
- Increasingly, people like Robert Kosara (at Tableau) are treating this as an empirical question for research.

Kieran Healy has a chapter on the good and the bad in data visualization, which I'd strongly encourage you to read: http://socviz.co/lookatdata.html

Here's an excerpt from the beginning: 
  
> Some data visualizations are better than others. This chapter discusses why that is. While it is tempting to simply start laying down the law about what works and what doesn’t, the process of making a really good or really useful graph cannot be boiled down to a list of simple rules to be followed without exception in all circumstances. The graphs you make are meant to be looked at by someone. The effectiveness of any particular graph is not just a matter of how it looks in the abstract, but also a question of who is looking at it, and why.   

### Plot 1:  

```{r}

```

### Plot 2:  

```{r}

```

### Plot 3:  

```{r}

```

Hints

1.5.
Example questions are:
- Why are there more diamonds at who carats and common fraction of carats?
- Why are there no diamonds bigger than 3 carats?
Clusters of similar values suggest that subgroups exist in your data. Your question can also be about understanding the subgroups. 

2.1.
Try to play around with geom_freqpoly and geom_boxplot. Which one seems better?

2.3. 
Use `geom_point`